# 1950s data from NHGIS

Pulling in data from 1950

```{r, message=F, warning=F}

pop_1950 <- read.csv("../../data/time-series/nhgis0006_csv/nhgis0006_ds82_1950_tract.csv")

#View(pop_1950)

pop_1950 %>% 
  filter(STATE == "Alabama")

```

## 1950 Shape files

```{r, message=F, warning=F}

tracts_1950 <- st_read(
  "../../data/time-series/nhgis0006_shape/nhgis0006_shapefile_tl2000_us_tract_1950/US_tract_1950.shp"
)

head(tracts_1950)
plot(tracts_1950["geometry"])

mobile_1950 <- tracts_1950 %>%
  dplyr::filter(
    NHGISST == "010",
    NHGISCTY == "0970"
  )

unique(tracts_1950$NHGISST)

unique(tracts_1950$NHGISCTY)

# GIS join for more robust analysis
## NHGIS IDs are hierarchical: G0100970XXXX
mobile_1950 <- tracts_1950 %>%
  dplyr::filter(grepl("^G0100970", GISJOIN))

# to overlay with modern data
mobile_1950 <- st_transform(mobile_1950, 4326)
```

Note: 1950 tract geography is inconsistent with later decades and sometimes incomplete (tracts didn’t cover all areas yet).

# 1960s data from NHGIS

```{r, message=F, warning=F}

pop_1960 <- read.csv("../../data/time-series/nhgis0007_csv/nhgis0007_ds92_1960_tract.csv")

#View(pop_1960)

pop_1960 %>% 
  filter(STATE == "Alabama")

```


## 1960 Shape files

```{r, message=F, warning=F}

tracts_1960 <- st_read(
  "../../data/time-series/nhgis0007_shape/nhgis0007_shapefile_tl2000_us_tract_1960/US_tract_1960.shp"
)

head(tracts_1960)
plot(tracts_1960["geometry"])

mobile_1960 <- tracts_1960 %>%
  dplyr::filter(
    NHGISST == "010",
    NHGISCTY == "0970"
  )

unique(tracts_1960$NHGISST)

unique(tracts_1960$NHGISCTY)

# GIS join for more robust analysis
## NHGIS IDs are hierarchical: G0100970XXXX
mobile_1960 <- tracts_1960 %>%
  dplyr::filter(grepl("^G0100970", GISJOIN))

# to overlay with modern data
mobile_1960 <- st_transform(mobile_1960, 4326)
```


# 1970 data

```{r, message=F, warning=F}


```



# 1970 Shape files

```{r, message=F, warning=F}

tracts_1970 <- st_read(
  "../../data/time-series/nhgis0007_shape/nhgis0007_shapefile_tl2000_us_tract_1970/US_tract_1970.shp"
)

head(tracts_1970)
plot(tracts_1970["geometry"])

mobile_1970 <- tracts_1970 %>%
  dplyr::filter(
    NHGISST == "010",
    NHGISCTY == "0970"
  )

unique(tracts_1970$NHGISST)

unique(tracts_1970$NHGISCTY)

# GIS join for more robust analysis
## NHGIS IDs are hierarchical: G0100970XXXX
mobile_1970 <- tracts_1970 %>%
  dplyr::filter(grepl("^G0100970", GISJOIN))

# to overlay with modern data
mobile_1970 <- st_transform(mobile_1970, 4326)
```



# Building a time series

Get 2020 tract geometries

```{r, message=F, warning=F}
tracts_2020 <- get_decennial(
  geography = "tract",
  state = "AL",
  county = "Mobile",
  variables = c(total = "P1_001N"),
  geometry = TRUE,
  year = 2020
) %>%
  select(GEOID, geometry)

nrow(tracts_2020)
```

Load 1960 NHGIS shapefile + data

```{r, message=F, warning=F}
tracts_1960 <- st_read("../../data/time-series/nhgis0007_shape/nhgis0007_shapefile_tl2000_us_tract_1960/US_tract_1960.shp") %>%
  filter(NHGISST == "010", NHGISCTY == "0970")

nhgis_1960 <- read.csv("../../data/time-series/nhgis0007_csv/nhgis0007_ds92_1960_tract.csv")

tracts_1960 <- tracts_1960 %>%
  left_join(nhgis_1960, by = "GISJOIN")

```

Harmonize CRS

** update this so that 1960 is mapped to 2010
** update this so that 1970 is mapped to 2020

```{r, message=F, warning=F}
tracts_1960 <- st_transform(tracts_1960, st_crs(tracts_2020))
```

Areal interpolation

```{r, message=F, warning=F}
#install.packages("areal")
library(areal)

tracts_1960_clean <- tracts_1960 %>%
  sf::st_make_valid() %>%
  sf::st_transform(st_crs(tracts_2020)) %>%
  dplyr::mutate(
    total = B79001,
    white = B79002,
    black = B79003
  ) %>%
  dplyr::select(total, white, black, geometry)

tracts_1960_clean
```

```{r, message=F, warning=F}
intersections <- st_intersection(
  tracts_1960_clean,
  tracts_2020
)
```



```{r, message=F, warning=F}
tracts_1960_clean <- st_make_valid(tracts_1960_clean)
tracts_2020 <- st_make_valid(tracts_2020)

any(!st_is_valid(tracts_1960_clean))
any(!st_is_valid(tracts_2020))

tracts_1960_clean <- st_simplify(
  tracts_1960_clean,
  dTolerance = 0,
  preserveTopology = TRUE
)

intersections <- st_intersection(
  tracts_1960_clean,
  tracts_2020
)

intersections

```


```{r, message=F, warning=F}

intersections <- sf::st_make_valid(intersections)

sf::sf_use_s2(FALSE)

intersections$int_area <- sf::st_area(intersections)

intersections$int_area <- as.numeric(intersections$int_area)
```

compute source tract total area


```{r, message=F, warning=F}
tract_area <- sf::st_area(tracts_1960_clean)
tract_area <- as.numeric(tract_area)

tract_ids <- tracts_1960_clean$GISJOIN
names(tract_area) <- tract_ids
```

compute weights for intersections

```{r, eval=F, message=F, warning=F}
intersections <- intersections %>%
  mutate(
    source_area = tract_area[GISJOIN],
    weight = int_area / source_area
  )

```

---
