---
title: "Neighborhood Data View: Hough, Cleveland, OH"
subtitle: "Initial data reporting for the PIT UN Project"
author: 
- "Quantitative Histories Workshop"
- "Nathan Alexander, Bayow Onabajo, Lyrric Jackson"
output: 
  html_document:
    theme: journal
    toc: true
    toc_float: true
    mathjax: default
    self_contained: true
  revealjs::revealjs_presentation:
    self_contained: true
    mathjax: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, include=F, echo=F, warning = FALSE, message = FALSE}
# options(tigris_use_cache = TRUE)
# Load necessary libraries
library(tidycensus)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(here)
here::i_am("cleveland-overview.Rmd")
```

# Overview

We provide a preliminary analysis of factors related to race and ethnicity and the Hough neighborhood in Cleveland, Ohio. We use data from the U.S. Census Bureau and local data collected from Cleveland Open Data ^[https://data.clevelandohio.gov/]. This information will be used to inform work regarding integrated data systems.

# Demography

## Cleveland Open Data

County and neighborhood information.

- SPA: Statistical Planning Areas

- SPANM: Name of the Statistical Planning Area

- SQ_MILES: The area of the Statistical Planning Area in square miles

```{r, results=T, warning = FALSE, message = FALSE}
# neighborhoods in Cleveland from open data site
neighborhoods <- 
  read.csv("../data/cleveland_neighborhoods.csv")
neighborhoods %>% 
  select(SPA, SPANM, SQ_MILES)
```

## Census data

Variable list of acs 5-year estimates for initial analysis.

```{r, results=F, warning = FALSE, message = FALSE}
# gathering relevant variables
vars <- c(
  total_population = 
    "B03002_001", # Total population (universe for race/ethnicity)
  
  # race-ethnicity vars
  white_alone = "B03002_003", # White alone
  black_alone = "B03002_004", # Black or African American alone
  native_alone = "B03002_005", # American Indian and Alaska Native alone
  asian_alone = "B03002_006", # Asian alone
  nhpi_alone = "B03002_007", # Native Hawaiian, Pacific Islander alone
  other_alone = "B03002_008", # Some other race alone
  two_or_more = "B03002_009", # Two or more races
  hila_any = "B03002_012", # Hispanic or Latino (of any race)
  
  # housing and income vars
  below_10k = "B19001_002", # households with income less than $10K
  median_home_value = "B25077_001", # median
  public_housing = "B25003_002", # subsidized rental units
  public_assistance = "B19058_001", # households receiving public assistance
  median_household_income = "B19013_001", # median household income
  bachelor = "B15003_017", # population over 25 with a Bachelor's degree or higher
  
  # demographic vars
  male_population = "B01001_002", # male population
  female_population = "B01001_026", # female population
  median_age = "B01002_001", # median age
  employed = "B23025_002", # employed population (age 16+)
  unemployed = "B23025_003", # unemployed population (age 16+)
  children_under_18 = "B09001_001" # total number of children under 18 years
  )
```

### Census database for Cuyahoga County by year

Create data frames for each year from 2012 to 2022.

```{r, results=F, warning = FALSE, message = FALSE}
# creating a function to gather census data based on var list
get_census_data <- function(year) {
  census_data <- get_acs(
    geography = "tract",
    variables = vars,
    year = year,
    state = "OH",
    county = "Cuyahoga", # Cuyahoga covers Cleveland area
    output = "wide",
    geometry = TRUE,      # spatial data for mapping
    survey = "acs5"       # explicitly use 5-year ACS estimates
  )
  return(census_data)
}
```

```{r, results=F, warning = FALSE, message = FALSE}
# creating a list of years from 2012 to 2022
years <- 2012:2022

# using the lapply function to create a list of data frames
df_list <- 
  lapply(years, get_census_data)

# assigning each data frame to a variable named df10, df11, ..., df22
for (i in seq_along(years)) {
  assign(paste0("df", substr(years[i], 3, 4)), 
         df_list[[i]])
}

# printing the first few rows of each data frame to verify
for (year in years) {
  cat("\nData for", year, ":\n")
  print(head(get(paste0("df", substr(year, 3, 4)))))
}
```

### Single Indicator Views

```{r}
ggplot(df22) +
  geom_sf(aes(fill = black_aloneE)) +
  scale_fill_viridis_c() +
  labs(title = "Black Population by Tract, 2022")
```

```{r}
ggplot(df22) +
  geom_sf(aes(fill = white_aloneE)) +
  scale_fill_viridis_c() +
  labs(title = "White Population by Tract, 2022")
```

### Binary Indicators

| Indicator  | ACS Variable   | Description                                    |
|:-----------|:--------------|:-----------------------------------------------|
| white_10k  | B19001A_002   | White alone household, income < $10,000^[https://pearson.socialexplorer.com/data/ACS2019_5yr/metadata/?ds=ACS19_5yr&var=B19001A002]        |
| black_10k  | B19001B_002   | Black alone household, income < $10,000^[https://pearson.socialexplorer.com/data/ACS2019_5yr/metadata/?ds=ACS19_5yr&var=B19001B002]        |

```{r}

df22 <- df22 %>%
  filter(!is.na(white_10kE), !is.na(white_aloneE), white_aloneE > 0,
         !is.na(black_10kE), !is.na(black_aloneE), black_aloneE > 0) %>%
  mutate(
    prop_white_10k = white_10kE / white_aloneE,
    prop_black_10k = black_10kE / black_aloneE
  )

df22 <- df22 %>%
  mutate(
    white_10kE = as.numeric(white_10kE),
    black_10kE = as.numeric(black_10kE),
    white_aloneE = as.numeric(white_aloneE),
    black_aloneE = as.numeric(black_aloneE)
  ) %>%
  mutate(
    prop_white_10k = white_10kE / white_aloneE,
    prop_black_10k = black_10kE / black_aloneE
  )

```


```{r}
ggplot(df22, aes(x = prop_white_10k)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.6, na.rm = TRUE) +
  labs(title = "Proportion of White Households Earning < $10k", x = "Proportion", y = "Number of Tracts") +
  theme_minimal()

ggplot(df22, aes(x = prop_black_10k)) +
  geom_histogram(bins = 30, fill = "red", alpha = 0.6, na.rm = TRUE) +
  labs(title = "Proportion of Black Households Earning < $10k", x = "Proportion", y = "Number of Tracts") +
  theme_minimal()


```


# State Violence

Examining fatal police shooting data for Cuyahoga County, which covers the Cleveland area.

```{r}
fatal <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/v2/fatal-police-shootings-data.csv") %>% 
  as_tibble()
head(fatal, n=3)
```

## Data for Cuyahoga County, Ohio

```{r}
fatal_cuyahoga <- fatal %>% 
  filter(county == "Cuyahoga") %>% 
  select(name,
         date,
         age, 
         gender, 
         race, 
         body_camera, 
         threat_type, 
         armed_with,
         flee_status,
         state,
         city,
         county,
         id)

# change vars to appropriate formats
fatal_cuyahoga$date <- as.Date(fatal_cuyahoga$date) # check/change to date format
fatal_cuyahoga$age <- as.numeric(fatal_cuyahoga$age)
fatal_cuyahoga$body_camera <- as.logical(fatal_cuyahoga$body_camera)

fatal_cuyahoga_year <- format(fatal_cuyahoga$date, format="20%y") 
fatal_cuyahoga$year <- fatal_cuyahoga_year # add a year column to the df
fatal_cuyahoga <- fatal_cuyahoga %>% 
  relocate(county, city, name, date)
fatal_cuyahoga
```

```{r}
fatal_cuyahoga %>%
  count(year) %>%
  ggplot(aes(x = factor(year), y = n)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Fatal police interactions with police officers by year - Cuyahoga",
    x = "Year",
    y = "Count"
  ) +
  theme_minimal()
```


```{r}


```



