---
title: "Historical Census Tracts from DC Open Data"
author: 
  - "Nathan Alexander, PhD"
  - "Center for Applied Data Science and Analytics"
  - "Howard University"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
    always_allow_html: true
    self_contained: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=T, warning=F, results=F, message=F}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(here)
here::i_am("dc_historical_census_tracts.Rmd")
```

# DATA

Files on the Census tracts from various historical periods are uploaded from DC Open Data. Additional data files can be accessed from the DC Open Data webiste [here](https://opendata.dc.gov/) and include KML, GeoJSON, and shapefies.

## Historical Census Tracts for DC

We load the data for historical tracts for DC to utilize and combine with other data.

### Year 1940

```{r}
dc.tracts.1940 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_1940.csv")
head(dc.tracts.1940)
```

### Year 1950

```{r}
dc.tracts.1950 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_1950.csv")
head(dc.tracts.1950)
```

### Year 1960

```{r}
dc.tracts.1960 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_1960.csv")
head(dc.tracts.1960)
```

### Year 1970

```{r}
dc.tracts.1970 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_1970.csv")
head(dc.tracts.1970)
```

### Year 1980

```{r}
dc.tracts.1980 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_1980.csv")
head(dc.tracts.1980)
```

### Year 1990

Data for the year 1990 was not available.

### Year 2000

```{r}
dc.tracts.2000 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_2000.csv")
head(dc.tracts.2000)
```

### Year 2010

```{r}
dc.tracts.2010 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_2010.csv")
head(dc.tracts.2010)
```

### Year 2020

```{r}
dc.tracts.2020 <- read.csv("../data/dc_historical_tracts/Census_Tracts_in_2020.csv")
head(dc.tracts.2020)
```

