---
title: "Sample research report"
author: 
- "Nathan Alexander"
- "Center for Applied Data Science and Analytics (CADSA)"
- "Howard University"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
  toc: true
  toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)
library(ggplot2)
```


```{r}
# Set seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Create sample Census tract identifiers
# Format: State code (2 digits) + County code (3 digits) + Tract number (6 digits)
state_code <- sample(1:50, n, replace = TRUE)
county_code <- sample(1:999, n, replace = TRUE)
tract_number <- sample(1:999999, n, replace = TRUE)

census_tract <- sprintf("%02d%03d%06d", state_code, county_code, tract_number)

# Create sample data frame
data <- data.frame(
  census_tract = census_tract,
  total_population = round(runif(n, 1000, 50000)),
  black_population = round(runif(n, 50, 15000)),
  lower_income = round(runif(n, 10, 1000)),
  median_home_value = round(runif(n, 100000, 1000000)),
  public_housing = round(runif(n, 0, 500)),
  public_assistance = round(runif(n, 10, 1000)),
  median_household_income = round(runif(n, 30000, 150000)),
  bachelor = round(runif(n, 100, 10000)),
  white_population = round(runif(n, 100, 30000)),
  hispanic_population = round(runif(n, 50, 10000)),
  male_population = round(runif(n, 500, 25000)),
  female_population = round(runif(n, 500, 25000)),
  median_age = round(runif(n, 25, 50), 1),
  employed = round(runif(n, 500, 20000)),
  unemployed = round(runif(n, 10, 2000)),
  children_under_18 = round(runif(n, 100, 10000))
  )

# Display the first few rows of the dataset
head(data)

# Summary statistics
summary(data)
```

-   What is the total population across all census tracts in the sample?

```{r}
total_pop <- sum(data$total_population)
cat("The total population across all census tracts is", total_pop)
```
-   How does the racial composition vary across census tracts?

```{r}
data %>%
  summarise(
    white_pct = mean(white_population / total_population * 100),
    black_pct = mean(black_population / total_population * 100),
    hispanic_pct = mean(hispanic_population / total_population * 100)
  )
```
-   What is the average median household income across all tracts?

```{r}
avg_median_income <- mean(data$median_household_income)
cat("The average median household income is $", round(avg_median_income, 2))

```

-   Is there a correlation between median home value and median household income?

```{r}
correlation <- cor(data$median_home_value, data$median_household_income)
cat("The correlation between median home value and median household income is", round(correlation, 3))

```

-   How does the proportion of the population with a bachelor's degree or higher vary across tracts?

```{r}
ggplot(data, aes(x = bachelor / total_population * 100)) +
  geom_histogram(bins = 20) +
  labs(x = "Percentage with Bachelor's degree", y = "Count")

```

-   What is the average unemployment rate across the sampled tracts?

```{r}
avg_unemployment_rate <- mean(data$unemployed / (data$employed + data$unemployed) * 100)
cat("The average unemployment rate is", round(avg_unemployment_rate, 2), "%")

```

-   Is there a relationship between the percentage of households receiving public assistance and median household income?

```{r}
ggplot(data, aes(x = median_household_income, y = public_assistance / total_population * 100)) +
  geom_point() +
  labs(x = "Median Household Income", y = "Percentage Receiving Public Assistance")

```

-   How does the ratio of male to female population vary across tracts?

```{r}
data %>%
  mutate(male_female_ratio = male_population / female_population) %>%
  summarise(
    min_ratio = min(male_female_ratio),
    max_ratio = max(male_female_ratio),
    mean_ratio = mean(male_female_ratio)
  )

```

-   What is the average median age across all tracts?

```{r}
avg_median_age <- mean(data$median_age)
cat("The average median age across all tracts is", round(avg_median_age, 1), "years")

```

-   Is there a correlation between the number of children under 18 and the total population in each tract?

```{r}
correlation <- cor(data$children_under_18, data$total_population)
cat("The correlation between children under 18 and total population is", round(correlation, 3))

```

