---
title: "Census research report - CADSA RDC Fellowship"
author: 
- "Nathan Alexander"
- "Center for Applied Data Science and Analytics (CADSA)"
- "Howard University"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
  toc: true
  toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Load necessary libraries
library(tidycensus)
library(dplyr)
library(ggplot2)
```

```{r}
# libraries
library(tidycensus)
library(dplyr)
library(ggplot2)
# options(tigris_use_cache = TRUE)

# variables
vars <- c(
  total_population = "B01001_001",         # Total Population
  black_population = "B03002_004",         # Black or African American Population
  lower_income = "B19001_002",             # Households with income less than $10,000
  median_home_value = "B25077_001",        # Median Home Value
  public_housing = "B25003_002",           # Occupied housing units that are renter-occupied and subsidized
  public_assistance = "B19058_001",        # Households Receiving Public Assistance
  median_household_income = "B19013_001",  # Median Household Income
  bachelor = "B15003_017",                 # Population over 25 with a Bachelor's degree or higher
  white_population = "B03002_003",         # White Population
  hispanic_population = "B03002_012",      # Hispanic or Latino Population
  male_population = "B01001_002",          # Male Population
  female_population = "B01001_026",        # Female Population
  median_age = "B01002_001",               # Median Age
  employed = "B23025_002",                 # Employed population (age 16+)
  unemployed = "B23025_003",               # Unemployed population (age 16+)
  children_under_18 = "B09001_001"         # Total number of own children under 18 years
  )

get_census_data <- function(year) {
  census_data <- get_acs(
    geography = "tract",
    variables = vars,
    year = year,
    state = "DC",
    output = "wide"
  )
}  
```

```{r}
library(tidycensus)

get_census_data <- function(year) {
  census_data <- get_acs(
    geography = "tract",
    variables = vars,
    year = year,
    state = "DC",
    output = "wide"
  )
  return(census_data)
}

# Create a list of years from 2012 to 2022
years <- 2012:2022

# Use lapply to create a list of data frames
df_list <- lapply(years, get_census_data)

# Assign each data frame to a variable named df10, df11, ..., df22
for (i in seq_along(years)) {
  assign(paste0("df", substr(years[i], 3, 4)), df_list[[i]])
}

# Print the first few rows of each data frame to verify
for (year in years) {
  cat("\nData for", year, ":\n")
  print(head(get(paste0("df", substr(year, 3, 4)))))
}

```

```{r}
# view data
head(df21)
head(df14)

# summary stats
summary(df21)
```


-   What is the total population across all census tracts in the sample?

```{r}
total_pop <- sum(df22$total_populationE)
cat("The total population across all census tracts is", total_pop)

total_pop <- sum(df22$total_populationE, na.rm = TRUE)
total_pop_moe <- sqrt(sum(df22$total_populationM^2, na.rm = TRUE))
cat("The total population across all census tracts is:", total_pop, 
    "±", round(total_pop_moe, 2), "(margin of error)")
```

-   How does the racial composition vary across census tracts?

```{r}
df22 %>%
  summarise(
    white_pct = mean(white_populationE / total_populationE * 100),
    black_pct = mean(black_populationE / total_populationE * 100),
    hispanic_pct = mean(hispanic_populationE / total_populationE * 100)
  )

df22 %>%
  summarise(
    white_pct = mean(white_populationE / total_populationE * 100, na.rm = TRUE),
    white_pct_moe = sqrt(mean((white_populationM / total_populationE * 100)^2, na.rm = TRUE)),
    black_pct = mean(black_populationE / total_populationE * 100, na.rm = TRUE),
    black_pct_moe = sqrt(mean((black_populationM / total_populationE * 100)^2, na.rm = TRUE)),
    hispanic_pct = mean(hispanic_populationE / total_populationE * 100, na.rm = TRUE),
    hispanic_pct_moe = sqrt(mean((hispanic_populationM / total_populationE * 100)^2, na.rm = TRUE))
  )

```

-   What is the average median household income across all tracts?

```{r}
avg_median_income <- mean(df22$median_household_incomeE, na.rm=T)
cat("The average median household income is $", round(avg_median_income, 2))

avg_median_income <- mean(df22$median_household_incomeE, na.rm = TRUE)
avg_median_income_moe <- sqrt(mean(df22$median_household_incomeM^2, na.rm = TRUE))
cat("The average median household income is: $", round(avg_median_income, 2),
    "±", round(avg_median_income_moe, 2), "(margin of error)")

```

-   Is there a correlation between median home value and median household income?

```{r}
correlation <- cor(df22$median_home_valueE, df22$median_household_incomeE, use = "complete.obs")
cat("The correlation between median home value and median household income is", round(correlation, 3))

correlation <- cor(df22$median_home_valueE, df22$median_household_incomeE, use = "complete.obs")
cat("The correlation between median home value and median household income is:", round(correlation, 3))

```

-   How does the proportion of the population with a bachelor's degree or higher vary across tracts?

```{r}
ggplot(df22, aes(x = bachelorE / total_populationE * 100)) +
  geom_histogram(bins = 20) +
  labs(x = "Percentage with Bachelor's degree", y = "Count")

df22 %>%
  mutate(bachelor_pct = bachelorE / total_populationE * 100,
         bachelor_pct_moe = sqrt((bachelorM / total_populationE * 100)^2 + 
                                 (bachelorE * total_populationM / total_populationE^2 * 100)^2)) %>%
  ggplot(aes(x = bachelor_pct)) +
  geom_histogram(bins = 20) +
  labs(x = "Percentage with Bachelor's degree", y = "Count")

```
-   What is the average unemployment rate across the sampled tracts?

```{r}
avg_unemployment_rate <- mean(df22$unemployedE / (df22$employedE + df22$unemployedE) * 100)
cat("The average unemployment rate is", round(avg_unemployment_rate, 2), "%")

df22 %>%
  mutate(unemployment_rate = unemployedE / (employedE + unemployedE) * 100,
         unemployment_rate_moe = sqrt((unemployedM / (employedE + unemployedE) * 100)^2 +
                                      (unemployedE * employedM / (employedE + unemployedE)^2 * 100)^2)) %>%
  summarise(avg_unemployment_rate = mean(unemployment_rate, na.rm = TRUE),
            avg_unemployment_rate_moe = sqrt(mean(unemployment_rate_moe^2, na.rm = TRUE))) %>%
  print()

```

Relationship between public assistance and median household income

```{r}
df22 %>%
  mutate(public_assistance_pct = public_assistanceE / total_populationE * 100) %>%
  ggplot(aes(x = median_household_incomeE, y = public_assistance_pct)) +
  geom_point() +
  labs(x = "Median Household Income", y = "Percentage Receiving Public Assistance")
```

Male to female population ratio variation

```{r}

df22 %>%
  mutate(male_female_ratio = male_populationE / female_populationE,
         male_female_ratio_moe = sqrt((male_populationM / female_populationE)^2 +
                                      (male_populationE * female_populationM / female_populationE^2)^2)) %>%
  summarise(
    min_ratio = min(male_female_ratio, na.rm = TRUE),
    max_ratio = max(male_female_ratio, na.rm = TRUE),
    mean_ratio = mean(male_female_ratio, na.rm = TRUE),
    mean_ratio_moe = sqrt(mean(male_female_ratio_moe^2, na.rm = TRUE))
  )

```

Average median age

```{r}
avg_median_age <- mean(df22$median_ageE, na.rm = TRUE)
avg_median_age_moe <- sqrt(mean(df22$median_ageM^2, na.rm = TRUE))
cat("The average median age across all tracts is:", round(avg_median_age, 1), 
    "±", round(avg_median_age_moe, 1), "years (margin of error)")
```

Correlation between children under 18 and total population

```{r}
correlation <- cor(df22$children_under_18E, df22$total_populationE, use = "complete.obs")
cat("The correlation between children under 18 and total population is:", round(correlation, 3))
```


```{r}

library(tidyverse)
library(knitr)

# Create a list of all dataframes
df_list <- list(df12, df13, df14, df15, df16, df17, df18, df19, df20, df21, df22)

# Extract year from each dataframe name
years <- 2012:2022

# Combine all dataframes into one, adding a year column
combined_df <- map2_dfr(df_list, years, ~mutate(.x, year = .y))

# Calculate the mean median household income for each year
summary_df <- combined_df %>%
  group_by(year) %>%
  summarise(
    mean_income = mean(median_household_incomeE, na.rm = TRUE),
    se_income = sd(median_household_incomeE, na.rm = TRUE) / sqrt(n())
  )

# Create the plot
ggplot(summary_df, aes(x = year, y = mean_income)) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  theme_minimal() +
  labs(title = "Trend in Median Household Income in DC (2012-2022)",
       x = "Year",
       y = "Median Household Income ($)") +
  scale_x_continuous(breaks = 2012:2022) +
  scale_y_continuous(labels = scales::dollar_format())

# Create summary table
summary_table <- combined_df %>%
  group_by(year) %>%
  summarise(
    mean_income = mean(median_household_incomeE, na.rm = TRUE),
    median_income = median(median_household_incomeE, na.rm = TRUE),
    sd_income = sd(median_household_incomeE, na.rm = TRUE),
    min_income = min(median_household_incomeE, na.rm = TRUE),
    max_income = max(median_household_incomeE, na.rm = TRUE)
  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

# Display the table
kable(summary_table, 
      col.names = c("Year", "Mean", "Median", "Std Dev", "Min", "Max"),
      caption = "Summary of Median Household Income by Year (2012-2022)")


```

```{r}
library(tidyverse)
library(ggplot2)
library(knitr)

# Create a list of all dataframes
df_list <- list(df12, df13, df14, df15, df16, df17, df18, df19, df20, df21, df22)

# Extract year from each dataframe name
years <- 2012:2022

# Combine all dataframes into one, adding a year column
combined_df <- map2_dfr(df_list, years, ~mutate(.x, year = .y))

# Function to create summary table
create_summary_table <- function(data, variable) {
  data %>%
    group_by(year) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      median = median({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      min = min({{variable}}, na.rm = TRUE),
      max = max({{variable}}, na.rm = TRUE)
    ) %>%
    mutate(across(where(is.numeric), ~round(., 2)))
}

# Function to create smooth line plot
create_smooth_plot <- function(data, variable, title, y_label) {
  ggplot(data, aes(x = year, y = {{variable}})) +
    geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
    theme_minimal() +
    labs(title = title, x = "Year", y = y_label) +
    scale_x_continuous(breaks = 2012:2022)
}

# Total Population
total_pop_summary <- create_summary_table(combined_df, total_populationE)
total_pop_plot <- create_smooth_plot(combined_df, total_populationE, 
                                     "Trend in Total Population (2012-2022)", 
                                     "Total Population")

# Median Home Value
home_value_summary <- create_summary_table(combined_df, median_home_valueE)
home_value_plot <- create_smooth_plot(combined_df, median_home_valueE, 
                                      "Trend in Median Home Value (2012-2022)", 
                                      "Median Home Value ($)")

# Display tables
cat("Summary of Total Population by Year (2012-2022)\n")
print(kable(total_pop_summary, 
            col.names = c("Year", "Mean", "Median", "Std Dev", "Min", "Max")))

cat("\nSummary of Median Home Value by Year (2012-2022)\n")
print(kable(home_value_summary, 
            col.names = c("Year", "Mean", "Median", "Std Dev", "Min", "Max")))

# Display plots
print(total_pop_plot)
print(home_value_plot)
```

```{r}
# Identify lowest 10% income neighborhoods
lowest_10_percent <- combined_df %>%
  group_by(year) %>%
  mutate(income_percentile = percent_rank(median_household_incomeE)) %>%
  filter(income_percentile <= 0.1)

# Function to create summary table
create_summary_table <- function(data, variable) {
  data %>%
    group_by(year) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      median = median({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      min = min({{variable}}, na.rm = TRUE),
      max = max({{variable}}, na.rm = TRUE)
    ) %>%
    mutate(across(where(is.numeric), ~round(., 2)))
}

# Function to create smooth line plot
create_smooth_plot <- function(data, variable, title, y_label) {
  ggplot(data, aes(x = year, y = {{variable}})) +
    geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
    theme_minimal() +
    labs(title = title, x = "Year", y = y_label) +
    scale_x_continuous(breaks = 2012:2022)
}

# Create plots and tables for all neighborhoods
all_income_plot <- create_smooth_plot(combined_df, median_household_incomeE, 
                                      "Median Household Income Trend (All Neighborhoods)", 
                                      "Median Household Income ($)")
all_pop_plot <- create_smooth_plot(combined_df, total_populationE, 
                                   "Total Population Trend (All Neighborhoods)", 
                                   "Total Population")
all_home_value_plot <- create_smooth_plot(combined_df, median_home_valueE, 
                                          "Median Home Value Trend (All Neighborhoods)", 
                                          "Median Home Value ($)")

all_income_summary <- create_summary_table(combined_df, median_household_incomeE)
all_pop_summary <- create_summary_table(combined_df, total_populationE)
all_home_value_summary <- create_summary_table(combined_df, median_home_valueE)

# Create plots and tables for lowest 10% income neighborhoods
low_income_plot <- create_smooth_plot(lowest_10_percent, median_household_incomeE, 
                                      "Median Household Income Trend (Lowest 10% Income Neighborhoods)", 
                                      "Median Household Income ($)")
low_pop_plot <- create_smooth_plot(lowest_10_percent, total_populationE, 
                                   "Total Population Trend (Lowest 10% Income Neighborhoods)", 
                                   "Total Population")
low_home_value_plot <- create_smooth_plot(lowest_10_percent, median_home_valueE, 
                                          "Median Home Value Trend (Lowest 10% Income Neighborhoods)", 
                                          "Median Home Value ($)")

low_income_summary <- create_summary_table(lowest_10_percent, median_household_incomeE)
low_pop_summary <- create_summary_table(lowest_10_percent, total_populationE)
low_home_value_summary <- create_summary_table(lowest_10_percent, median_home_valueE)

# Display plots side by side
library(gridExtra)
grid.arrange(all_income_plot, low_income_plot, ncol = 2)
grid.arrange(all_pop_plot, low_pop_plot, ncol = 2)
grid.arrange(all_home_value_plot, low_home_value_plot, ncol = 2)

# Display summary tables
print(kable(all_income_summary, caption = "Summary of Median Household Income (All Neighborhoods)"))
print(kable(low_income_summary, caption = "Summary of Median Household Income (Lowest 10% Income Neighborhoods)"))

print(kable(all_pop_summary, caption = "Summary of Total Population (All Neighborhoods)"))
print(kable(low_pop_summary, caption = "Summary of Total Population (Lowest 10% Income Neighborhoods)"))

print(kable(all_home_value_summary, caption = "Summary of Median Home Value (All Neighborhoods)"))
print(kable(low_home_value_summary, caption = "Summary of Median Home Value (Lowest 10% Income Neighborhoods)"))


```


```{r}
library(tidyverse)
library(knitr)
library(gridExtra)

# Create a list of all dataframes
df_list <- list(df12, df13, df14, df15, df16, df17, df18, df19, df20, df21, df22)

# Extract year from each dataframe name
years <- 2012:2022

# Combine all dataframes into one, adding a year column
combined_df <- map2_dfr(df_list, years, ~mutate(.x, year = .y))

# Identify lowest 10% Black population neighborhoods
lowest_10_percent_black <- combined_df %>%
  group_by(year) %>%
  mutate(black_population_percentile = percent_rank(black_populationE)) %>%
  filter(black_population_percentile <= 0.1)

# Function to create summary table
create_summary_table <- function(data, variable) {
  data %>%
    group_by(year) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      median = median({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      min = min({{variable}}, na.rm = TRUE),
      max = max({{variable}}, na.rm = TRUE)
    ) %>%
    mutate(across(where(is.numeric), ~round(., 2)))
}

# Function to create smooth line plot
create_smooth_plot <- function(data, variable, title, y_label) {
  ggplot(data, aes(x = year, y = {{variable}})) +
    geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
    theme_minimal() +
    labs(title = title, x = "Year", y = y_label) +
    scale_x_continuous(breaks = 2012:2022)
}

# Create plots and tables for all neighborhoods (Black population)
all_black_pop_plot <- create_smooth_plot(combined_df, black_populationE, 
                                         "Black Population Trend (All Neighborhoods)", 
                                         "Black Population")
all_black_pop_summary <- create_summary_table(combined_df, black_populationE)

# Create plots and tables for lowest 10% income neighborhoods (Black population)
low_black_pop_plot <- create_smooth_plot(lowest_10_percent_black, black_populationE, 
                                         "Black Population Trend (Lowest 10% Income Neighborhoods)", 
                                         "Black Population")
low_black_pop_summary <- create_summary_table(lowest_10_percent_black, black_populationE)

# Display plots side by side
grid.arrange(all_black_pop_plot, low_black_pop_plot, ncol = 2)

# Display summary tables
cat("Summary of Black Population (All Neighborhoods)\n")
print(kable(all_black_pop_summary, caption = "Summary of Black Population (All Neighborhoods)"))

cat("\nSummary of Black Population (Lowest 10% Income Neighborhoods)\n")
print(kable(low_black_pop_summary, caption = "Summary of Black Population (Lowest 10% Income Neighborhoods)"))


```

```{r}
library(tidyverse)
library(knitr)
library(gridExtra)

# Create a list of all dataframes
df_list <- list(df12, df13, df14, df15, df16, df17, df18, df19, df20, df21, df22)

# Extract year from each dataframe name
years <- 2012:2022

# Combine all dataframes into one, adding a year column
combined_df <- map2_dfr(df_list, years, ~mutate(.x, year = .y))

# Identify lowest 10% income neighborhoods
lowest_10_percent <- combined_df %>%
  group_by(year) %>%
  mutate(income_percentile = percent_rank(median_household_incomeE)) %>%
  filter(income_percentile <= 0.1)

# Create plots and tables for all neighborhoods (Public Housing)
all_public_housing_plot <- create_smooth_plot(combined_df, public_housingE, 
                                              "Public Housing Trend (All Neighborhoods)", 
                                              "Public Housing Units")
all_public_housing_summary <- create_summary_table(combined_df, public_housingE)

# Create plots and tables for lowest 10% income neighborhoods (Public Housing)
low_public_housing_plot <- create_smooth_plot(lowest_10_percent, public_housingE, 
                                              "Public Housing Trend (Lowest 10% Income Neighborhoods)", 
                                              "Public Housing Units")
low_public_housing_summary <- create_summary_table(lowest_10_percent, public_housingE)

# Display plots side by side
grid.arrange(all_public_housing_plot, low_public_housing_plot, ncol = 2)

# Display summary tables
cat("Summary of Public Housing Units (All Neighborhoods)\n")
print(kable(all_public_housing_summary, caption = "Summary of Public Housing Units (All Neighborhoods)"))

cat("\nSummary of Public Housing Units (Lowest 10% Income Neighborhoods)\n")
print(kable(low_public_housing_summary, caption = "Summary of Public Housing Units (Lowest 10% Income Neighborhoods)"))


```


```{r}
library(tidyverse)
library(knitr)
library(gridExtra)

# Create a list of all dataframes
df_list <- list(df12, df13, df14, df15, df16, df17, df18, df19, df20, df21, df22)

# Extract year from each dataframe name
years <- 2012:2022

# Combine all dataframes into one, adding a year column
combined_df <- map2_dfr(df_list, years, ~mutate(.x, year = .y))

# Identify lowest 10% income neighborhoods
lowest_10_percent <- combined_df %>%
  group_by(year) %>%
  mutate(income_percentile = percent_rank(median_household_incomeE)) %>%
  filter(income_percentile <= 0.1)

# Normalize proportions for all neighborhoods and lowest 10% income neighborhoods
combined_df <- combined_df %>%
  mutate(public_housing_proportion = public_housingE / total_populationE)

lowest_10_percent <- lowest_10_percent %>%
  mutate(public_housing_proportion = public_housingE / total_populationE)

# Create smooth line plot for both groups on the same plot
ggplot() +
  geom_smooth(data = combined_df, aes(x = year, y = public_housing_proportion), 
              method = "loess", color = "blue", fill = "lightblue", se = TRUE) +
  geom_smooth(data = lowest_10_percent, aes(x = year, y = public_housing_proportion), 
              method = "loess", color = "red", fill = "lightpink", se = TRUE) +
  theme_minimal() +
  labs(title = "Proportion of Public Housing Units (2012-2022)",
       x = "Year",
       y = "Proportion of Public Housing Units") +
  scale_x_continuous(breaks = years) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01))

# Create summary tables for proportions
all_public_housing_summary <- create_summary_table(combined_df, public_housing_proportion)
low_public_housing_summary <- create_summary_table(lowest_10_percent, public_housing_proportion)

# Display summary tables
cat("Summary of Proportions of Public Housing Units (All Neighborhoods)\n")
print(kable(all_public_housing_summary, caption = "Summary of Proportions of Public Housing Units (All Neighborhoods)"))

cat("\nSummary of Proportions of Public Housing Units (Lowest 10% Income Neighborhoods)\n")
print(kable(low_public_housing_summary, caption = "Summary of Proportions of Public Housing Units (Lowest 10% Income Neighborhoods)"))


```




