---
title: "Analyzing Decennial Census Data"
author: 
  - "Nathan Alexander, PhD"
  - "Center for Applied Data Science and Analytics"
  - "Howard University"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
    always_allow_html: true
    self_contained: true
editor_options:
  markdown:
    wrap: sentence
---

This file will support users in accessing the U.S. Decennial Census data. 

The code and data in this file is based on the 2025 workshop, "Working with Decennial Census Data in R" led by Prof. Kyle Walker (TCU); the slides for this workshop can be found [here](https://walker-data.com/umich-workshop-2025/decennial-census/#/title-slide) and the repository for the workshop can be found [here](https://github.com/walkerke/umich-workshop-2025).

Variable selection and targets have been prioritized for members of the Quantitative Histories Workshop and other quant-shop-users. Please note that you will need to have your census API key installed on your local machine.

```{r setup, include=T, warning=F, results=F, message=F}
knitr::opts_chunk$set(echo = TRUE)

pkgs <- c("tidycensus", "tidyverse", "mapview", "survey", "srvyr", "arcgislayers")
# install.packages(pkgs) # uncomment to install the packages

library(tidyverse)
library(tidycensus)
library(sf)
library(tidyverse)
library(ggplot2)
library(weights)
library(dplyr)
library(stringr)

options(tigris_use_cache = TRUE)
```

# Data

We will access the 2020 Census data using tidycensus.

## State-level population data

We will gather the 2020 state-level population data using the PL 94-171 Redistricting Data Summary File. This data was used to inform redistricting as a part of a U.S. constitutional mandate.

```{r, warning=F, message=F}
pop20 <- get_decennial(
  geography = "state",
  variables = "P1_001N",
  year = 2020
)

pop20
```

The table parameter obtains all related variables.

```{r, warning=F, message=F}
table_p2 <- get_decennial(
  geography = "state", 
  table = "P2", 
  year = 2020
)
```

```{r}
head(table_p2)
tail(table_p2)
```

## DC data

We will start with data on Washington, DC. 

Take note of the tibble dimensions.

```{r, message=F, warning=F}
dc_population <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  state = "DC",
  sumfile = "dhc",
  year = 2020
)

dc_population
```

## NC data

We will also collect data on North Carolina.

```{r, message=F, warning=F}
nc_population <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  state = "NC",
  sumfile = "dhc",
  year = 2020
)

nc_population
```

## State and County

Given the uniqueness of DC, we will focus on NC, specifically Mecklenburg County, NC.

From Walker, a census block is analogous to an urban block and it may vary by geography.

The block is the smallest geography that you can get in the decennial data.

```{r, message=F, warning=F}
nc_mecklenburg_blocks <- get_decennial(
  geography = "block",
  variables = "P1_001N",
  state = "NC",
  county = "Mecklenburg",
  sumfile = "dhc",
  year = 2020
)

nc_mecklenburg_blocks
```

# Variables

Next, we will identify our variables of interest.

```{r, message=F, warning=F}
vars <- load_variables(2020, "dhc")

View(vars) # view the variables, a new window will pop up
```

When the data frame loads in a new window, you can use the search function to identify variables.

Please note the different variable concepts and lables:

- P: person-level variables

- H: housing-level variables

- HCT: housing-level variables only available at census-tract level and up.

- PCT: person-level variables only available at the census tract level and up.

## Sumfiles

Take note of the `sumfile` that is used to access the data based on the variable name and concept.

`getdecennial()` defaults to `sumfile = pl` (Redistricting Data Summary file).

You will need to change the `sumfile = dhc`.

### Example: Black owner-occupied or renter-occupied housing

Taking a look at counts of owners and renters by Black or African American alone.

```{r, message=F, warning=F}
nc_housing_black <- get_decennial(
  geography = "county",
  state = "NC",
  variables = c(owner = "H10_004N",  # owner occupied
                renter = "H10_012N"), # renter occupied
  year = 2020,
  sumfile = "dhc",
  output = "wide"
)

nc_housing_black
```

From the example provided by Walker, I can look at the married and partnered by same sex.

```{r, message=F, warning=F}
nc_samesex <- get_decennial(
  geography = "county",
  state = "NC",
  variables = c(married = "DP1_0116P",
                partnered = "DP1_0118P"),
  year = 2020,
  sumfile = "dp",
  output = "wide"
)

nc_samesex
```

## Summary variables

We use the core-based statistical areas to take a look at summary variables.

We follow Walker and use an example with race-ethnicity breakdowns.

```{r, message=F, warning=F}
race_vars <- c(
  Hispanic = "P5_010N",
  White = "P5_003N",
  Black = "P5_004N",
  Native = "P5_005N",
  Asian = "P5_006N",
  HIPI = "P5_007N"
)

# Core-Based Statistical Areas (CBSA)
cbsa_race <- get_decennial(
  geography = "cbsa",
  variables = race_vars,
  summary_var = "P5_001N", 
  year = 2020,
  sumfile = "dhc"
)
```


The structure of the data determines whether we get a count or a proportion.

For example, in the code above, one options provided counts and the other proportions.

---

We can then arrange data to view the largest populated areas by race.

```{r, message=F, warning=F}
arrange(cbsa_race, desc(value)) %>% 
  head(n=15) # view the top ten areas
```

Now we need to view the data by proportions (percent) as to not use the raw data. 

```{r, message=F, warning=F}
cbsa_race_percent <- cbsa_race %>%
  mutate(percent = 100 * (value / summary_value)) %>% 
  select(NAME, variable, percent) 

cbsa_race_percent
```

Notice that the areas have changed.

---

## Group-wise analysis

We can view the largest proportions of groups by metro area.

```{r, message=F, warning=F}
largest_group <- cbsa_race_percent %>%
  group_by(NAME) %>% 
  filter(percent == max(percent))
```

We can analyze the largest proportion of Black communities across the country by metro.

```{r, message=F, warning=F}
largest_group %>% 
  arrange(desc(percent)) %>% 
  filter(variable == "Black")
```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```


```{r, message=F, warning=F}

```



```{r, message=F, warning=F}

```
