---
title: "Technical File: Examining Community Food Environments and Health Outcomes in Virginia"
author: 
- "Nathan Alexander, PhD"
- "Emmanuel Christian"
- "Howard University"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
    always_allow_html: true
    self_contained: true
editor_options:
  markdown:
    wrap: sentence
---

# Project setup

```{r setup, include=F, warning=F, message=F}
knitr::opts_chunk$set(echo = TRUE)
library(here)
here::i_am("technical-file-food-access-virginia.Rmd")

# install packages
#install.packages("tidycensus")
#install.packages("tidyverse")
#install.packages("mapview")
#install.packages("mapgl")
library(tidycensus)
library(dplyr)
library(ggplot2)
library(sf)
library(viridis)
library(scales)
library(readr)
```

# Food access data import

We import data from the [Food Access Research Atlas](https://www.ers.usda.gov/data-products/food-access-research-atlas).

```{r food-access-data, include=T, warning=F, message=F}
food.access <- read.csv("../../data/food-access/foodaccessdata2019.csv")
#head(food.access)
str(food.access)
```

# VA food access data subset

```{r va-food-access-data, include=T, warning=F, message=F}
va_food <- food.access %>%
  filter(State == "Virginia")

# convert to proper GEOID to merge with census data
va_food <- va_food %>%
  mutate(
    GEOID = sprintf("%011.0f", CensusTract)
  )
```

# Census data important

```{r census-data, include=T, warning=F, message=F}
va_census <- get_acs(
  geography = "tract",
  state = "VA",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001"
  ),
  year = 2020,
  survey = "acs5"
)

# reshape census data

va_census_wide <- va_census %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(
    names_from = variable,
    values_from = estimate
  )
```

# Merge data sets

```{r, include=T, warning=F, message=F}
va_merged <- va_food %>%
  left_join(va_census_wide, by = "GEOID")

# validate
nrow(va_food)
nrow(va_merged)

sum(is.na(va_merged$total_pop))
```

---

# Research Question

- General question: How does living in a community characterized as a food swamp, food desert, or food oasis affect the health outcomes of residents of these communities in Virginia Beach City County and Henrico County, VA?

- Focus question: How are tract-level food environment classifications associated with tract- or county-level health outcomes?

## Clean data

```{r, include=T, warning=F, message=F}
va_clean <- va_merged %>%
  mutate(
    across(where(is.character), ~na_if(., "NULL")),
    across(
      c(PovertyRate, MedianFamilyIncome, lalowihalfshare,
        lasnaphalfshare, lapophalfshare),
      readr::parse_number
    )
  )
```

## Study counties

```{r, include=T, warning=F, message=F}
va_study <- va_clean %>%
  filter(
    County %in% c("Virginia Beach city", "Henrico County")
  )
```

## Food environment classifications

```{r, include=T, warning=F, message=F}
va_study <- va_study %>%
  mutate(
    food_desert = LILATracts_1And10 == 1,
    food_oasis  = LILATracts_1And10 == 0 & LowIncomeTracts == 0
  )
```


We use a data driven threshold.

```{r, include=T, warning=F, message=F}
snap_cut   <- median(va_study$lasnaphalfshare, na.rm = TRUE)
access_cut <- median(va_study$lapophalfshare, na.rm = TRUE)

va_study <- va_study %>%
  mutate(
    food_swamp = LowIncomeTracts == 1 &
                 lasnaphalfshare > snap_cut &
                 lapophalfshare > access_cut
  )
```

Then we combine it into one variable.

```{r, include=T, warning=F, message=F}
va_study <- va_study %>%
  mutate(
    food_env = case_when(
      food_desert ~ "Food Desert",
      food_swamp  ~ "Food Swamp",
      food_oasis  ~ "Food Oasis",
      TRUE        ~ "Other"
    )
  )

```

## Import spatial data

```{r, include=F, warning=F, message=F}
va_tracts <- get_acs(
  geography = "tract",
  state = "VA",
  county = c("Virginia Beach city", "Henrico County"),
  variables = "B01003_001",
  geometry = TRUE
)

```

### Join data to the map

```{r, include=T, warning=F, message=F}
va_map <- va_tracts %>%
  select(GEOID, geometry) %>%
  left_join(va_study, by = "GEOID")

```

# Map 1: Food enviornment classifications

```{r, include=T, warning=F, message=F}
library(ggplot2)

ggplot(va_map) +
  geom_sf(aes(fill = food_env), color = NA) +
  scale_fill_manual(values = c(
    "Food Desert" = "red",
    "Food Swamp"  = "orange",
    "Food Oasis"  = "green",
    "Other"       = "grey80"
  )) +
  labs(
    title = "Food Environment Classification",
    subtitle = "Virginia Beach city & Henrico County",
    fill = "Category"
  ) +
  theme_minimal()

```


```{r, include=T, warning=F, message=F}


```




