---
title: "virtual_seminar_decennial_census"
author: "Bayowa"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)
```




# install.packages(c("tidycensus", "tidyverse", "mapview"))

# library(tidycensus)
# 
# census_api_key("YOUR KEY GOES HERE", install = TRUE)

```{r}
pop20 <- get_decennial(
  geography = "state",
  variables = "P1_001N",
  year = 2020
)

pop20
```


```{r}
table_p2 <- get_decennial(
  geography = "state", 
  table = "P2", 
  year = 2020
)

table_p2

```



```{r}
mo_population <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  state = "MO",
  sumfile = "dhc",
  year = 2020
)

mo_population
```


```{r}
stl_blocks <- get_decennial(
  geography = "block",
  variables = "P1_001N",
  state = "MO",
  county = "St. Louis city",
  sumfile = "dhc",
  year = 2020
)

stl_blocks
```

```{r}
vars <- load_variables(2020, "dhc")
 
View(vars)
```


```{r}
single_year_age <- get_decennial(
  geography = "state",
  table = "PCT12",
  year = 2020,
  sumfile = "dhc"
)


single_year_age
```


```{r}
single_year_age_wide <- get_decennial(
  geography = "state",
  table = "PCT12",
  year = 2020,
  sumfile = "dhc",
  output = "wide" 
)

single_year_age_wide
```
 
 
```{r}
fl_samesex <- get_decennial(
  geography = "county",
  state = "FL",
  variables = c(married = "DP1_0116P",
                partnered = "DP1_0118P"),
  year = 2020,
  sumfile = "dp",
  output = "wide"
)

fl_samesex
```





library(tidyverse)

tidyverse_logo()

```{r}
library(tidycensus)
library(tidyverse)

mo_population <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  year = 2020,
  state = "MO",
  sumfile = "dhc"
)


arrange(mo_population, value)
```


```{r}
arrange(mo_population, desc(value))

```


```{r}
below1000 <- filter(mo_population, value < 1000)

below1000

```


```{r}
below5000 <- filter(mo_population, value < 1000)

below5000

```


```{r}
countyquery <- filter(mo_population, value < 10000)

countyquery
```


```{r}
race_vars <- c(
  Hispanic = "P5_010N",
  White = "P5_003N",
  Black = "P5_004N",
  Native = "P5_005N",
  Asian = "P5_006N",
  HIPI = "P5_007N"
)

cbsa_race <- get_decennial(
  geography = "cbsa",
  variables = race_vars,
  summary_var = "P5_001N", 
  year = 2020,
  sumfile = "dhc"
)

cbsa_race
```


```{r}
cbsa_race_percent <- cbsa_race %>%
  mutate(percent = 100 * (value / summary_value)) %>% 
  select(NAME, variable, percent) 

cbsa_race_percent
```


```{r}
largest_group <- cbsa_race_percent %>%
  group_by(NAME) %>% 
  filter(percent == max(percent)) 

largest_group
```


#largest_group <- cbsa_race_percent %>%
  #group_by(NAME) %>% 
  #filter(percent == max(percent)) 

# Optionally, use `.by`: 
# largest_group <- cbsa_race_percent %>%
#   filter(percent == max(percent), .by = NAME) 

#largest_group


```{r}
cbsa_race_percent %>%
  group_by(variable) %>% 
  summarize(median_pct = median(percent, na.rm = TRUE)) 
```

# Spatial census data
```{r}
wv_over_65 <- get_decennial(
  geography = "tract",
  variables = "DP1_0024P",
  state = "WV",
  geometry = TRUE,
  sumfile = "dp",
  year = 2020
)


wv_over_65
```


```{r}
library(mapview)

mapview(wv_over_65)

```




```{r}
mapview(wv_over_65, zcol = "value")
```



```{r}
mapview(wv_over_65, zcol = "value",
        layer.name = "% age 65 and up<br>Census tracts in Iowa")
```



```{r}
library(viridisLite)

mapview(wv_over_65, zcol = "value",
        layer.name = "% age 65 and up<br>Census tracts in W. Virginia",
        col.regions = inferno(100))
```


```{r}
 library(htmlwidgets)
 
m2 <- mapview(wv_over_65, zcol = "value",
         layer.name = "% age 65 and up<br>Census tracts in W. Virginia",
         col.regions = inferno(100))
```


```{r}
saveWidget(m2@map, "wv_over_65.html")
```
 
 
# saveWidget(m1@map, "wv_over_65.html")

```{r}
mn_population_groups <- get_decennial(
  geography = "state",
  variables = "T01001_001N",
  state = "MN",
  year = 2020,
  sumfile = "ddhca",
  pop_group = "all",
  pop_group_label = TRUE
)


mn_population_groups
```
 

```{r}
available_groups <- get_pop_groups(2020, "ddhca")
available_groups
```



#code gets error
try({
get_decennial(
  geography = "county",
  variables = "T02001_001N",
  state = "MN",
  county = "Hennepin",
  pop_group = "1325",
  year = 2020,
  sumfile = "ddhca"
)
})

#check granular variables you can get
```{r}
check_ddhca_groups(
  geography = "county", 
  pop_group = "1325", 
  state = "MN", 
  county = "Hennepin"
)
```


```{r}
library(tidycensus)

hennepin_somali <- get_decennial(
  geography = "tract",
  variables = "T01001_001N",
  state = "MN",
  county = "Hennepin",
  year = 2020,
  sumfile = "ddhca",
  pop_group = "1325",
  pop_group_label = TRUE,
  geometry = TRUE
)
mapview(hennepin_somali, zcol = "value")
```
#gaps and sparse data are seen in the map above


```{r}
somali_dots <- as_dot_density(
  hennepin_somali,
  value = "value",
  values_per_dot = 25
)

mapview(somali_dots, cex = 0.01, layer.name = "Somali population<br>1 dot = 25 people",
        col.regions = "red", color = "red")
```
# mapview showing dots with each dot representing about 25 somalians in each region which is better than the zcol



```{r}
county_pop_10 <- get_decennial(
  geography = "county",
  variables = "P001001", 
  year = 2010,
  sumfile = "sf1"
)


county_pop_10

```


```{r}
county_pop_10_clean <- county_pop_10 %>%
  select(GEOID, value10 = value) 

county_pop_10_clean
```


```{r}
county_pop_20 <- get_decennial(
  geography = "county",
  variables = "P1_001N",
  year = 2020,
  sumfile = "dhc"
) %>%
  select(GEOID, NAME, value20 = value)
```


```{r}
county_joined <- county_pop_20 %>%
  left_join(county_pop_10_clean, by = "GEOID") 
county_joined
```


```{r}
county_change <- county_joined %>%
  mutate( 
    total_change = value20 - value10, 
    percent_change = 100 * (total_change / value10) 
  ) 


county_change
```


```{r}
filter(county_change, is.na(value10))
```
#counties that have a change in 2010: this illistrates why county census tract data isnt to good for time series analysis


